Parallel Information Set Generation for Kriegspiel

### 2012-09-15 sbac/paper.tex
ack
Addressed reviewer comments; fixed errors in figures.

%\documentclass[times, 10pt,twocolumn]{article}
%\documentclass[conference]{IEEEtran}
\documentclass[10pt, conference, compsocconf]{IEEEtran} 
%\usepackage{latex8}
%\usepackage{times}
\usepackage{url,hyperref}
\usepackage{subfigure}
\usepackage{tikz}
\usepackage{examplep}
\usepackage{moreverb}
\usepackage{algorithm}
\usepackage{boxedminipage}
\usepackage[noend]{algpseudocode}
\usepackage{listings}
\usepackage{amsmath,amsfonts,latexsym,graphicx}
\usepackage[tableposition=top,font=bf]{caption}
%\usepackage[tableposition=top,font=bf]{caption}

@@

\newcommand{\etal}{{\em et al.}\ }
\newcommand{\assign}{\leftarrow}
\newcommand{\eps}{\epsilon}
%\documentstyle[times,art10,twocolumn,latex8]{article}

@@

\begin{figure}
\begin{center}
\begin{verbatim}
1. d4 {(:)}
   a5 {(:)}
2. Bg5 {(:)}
   b6 {(:)}
3. Nc3 {(:Bd8)}
   c5 {(:)}
4. d5 {(:Bd8;Tc5)}
   Na6 {(:)}
5. d6 {(:)}
   f6 {(:Td6)}
6. e4 {(:)}
   fxg5 {(Xg5:Td6;Tg5)}
7. Bc4 {(:)}
   exd6 {(Xd6:Td6)}
8. Qd5 {(:)}
   Qc7 {(:)}
9. Qf7+ {(:)} 
   Kd8 {:g6,Kxf7,Ke7}
10. Qxf8++ {(1-0:)}
\end{verbatim}
\end{center}
\caption{Listing of a Kriegspiel game resulting in checkmate by white after 10 moves. The final, legal move for each player is shown followed by the list of moderator declarations and all moves that were attempted but disallowed before the legal move was accepted.}
\label{listing}
\vspace{-0.1in}
\end{figure}

@@

Figure~\ref{fig:figure1} shows the true state of the game after nine moves by
each player.  Figure~\ref{fig:figure2} shows White's perspective of the game at
that point. At move 9, Black attempts to block a
potential threat along the diagonal by moving a pawn to g6, then attempts to
escape or capture a threat at f7.  When both of these attempts fail, Black can
infer that his king is being threatened by a protected bishop or queen at f7.
The attempt to move to e7 is Black's way of finding out whether the threatening
piece is a queen or bishop.  When this move fails, black knows it is White's
queen and ultimately retreats to d8.  Note that the failed moves by White on
turns 3 and 4 are a result of being blocked by the pawn at e2.

The listing in Figure~\ref{listing} corresponds to the referee's view of the game and would be useful for
post-game analysis or commentary.  During the game, players would not have access to the full information.  The
transcript from White's perspective is shown in Figure~\ref{filteredlisting}.  Note that Black's actual moves have been
replaced with ``??'' and lists of Black's illegal moves have been replaced with the {\em number} of illegal attempted
moves.  For example, on move 9, Black attempted three illegal moves.
\begin{figure}
%\begin{minipage}[b]{0.35\linewidth}
%\centering
%\small
\begin{verbatim}
1. d4 {(:)}
   ?? {(:0)}
2. Bg5 {(:)}
   ?? {(:0)}
3. Nc3 {(:Bd8)}
   ??  {(:0)}
4. d5 {(:Bd8;Tc5)}
   ??  {(:0)}
5. d6 {(:)}
   ?? {(:0,Td6)}
6. e4 {(:)}
   fxg5 {(Xg5:Td6;Tg5)}
7. Bc4 {(:)}
   ?? {(Xd6:Td6)}
8. Qd5 {(:)}
   ?? {(:0)}
9. Qf7+ {(:)} 
   ?? {:0}
10. Qxf8++ {(1-0:)}
\end{verbatim}
%\end{verbatim}
%\end{minipage}
%\begin{minipage}[b]{0.25\linewidth}
%\small
%\centering
%\begin{verbatim}
%\end{minipage}
%\vspace{-0.1in}
\caption{Observations received by White over the course of the game.  White does not know Black's actual moves but hears moderator announcements related to captures, pawn tries, and the number of attempted moves by Black that fail.}
\label{filteredlisting}
\vspace{-0.2in}
\end{figure}

@@

Table~\ref{speedups} shows the performance results for our algorithm on a
randomly selected problem instance in which the size of the information set was
about 4.5 million.  This instance is not necessarily representative of all
games or states of the game, but it was not cherry-picked in any way. The table
shows that our implementation achieves good parallel scaling. The speedups are nearly
linear for up to 32 processors, and the speedups continue to increase even up
to 1,024 processors. In this particular instance, the sequential algorithm
takes almost 25 minutes to run.  However, when parallelized on 1,024
processors --- perfectly reasonable resource usage on modern hardware --- the
search takes less than three seconds.  This is a speedup of 576 times! In a
tournament where each player is allotted an average of five minutes per move,
an ISG operation that requires 25 minutes or more would be prohibitively
expensive.  On the other hand, if the operation can be done in under three
seconds, ISG can be used as a subroutine in the decision-making process
(see~\cite{parker05game}).

@@

\caption{Speedups for a randomly selected problem instance on 1--1,024
processors.  The input observations were from the 10th ply of the game tree and
the size of the information set in this instance is 4,495,121.}
\centering
\begin{tabular}{ccc}
{\bf Procs}	&	{\bf Time (s)} 	&	{\bf Speedup}\\
1	&	1524	&	1\\
2	&	812	&	1.88\\
4	&	407	&	3.74\\
8	&	208	&	7.32\\
16	&	102	&	14.8\\
32	&	53.2	&	28.7\\
64	&	28.0	&	54.5\\
128	&	14.9	&	102\\
256	&	7.70	&	197\\
512	&	4.59	&	331\\
1024	&	2.64	&	576\\
\end{tabular}

@@

We next study the variation in the efficiency of the parallelization of the
algorithm as size of the problem increases.  Here, we varied the number
of plies in the test instance from 8 to 12. We used the work stealing
load-balancing strategy.  
\begin{figure}[ht] 
\centering
\begin{minipage}{0.7\linewidth}
\includegraphics[width=\columnwidth]{plots/3depths.pdf}
\end{minipage}
\centering
\begin{minipage}{0.7\linewidth}
\includegraphics[width=\columnwidth]{plots/3depthspe.pdf}
\end{minipage}
\caption{(Top) Execution time for three different sizes of problem as the number of cores used increases.
(Bottom) Parallel Efficiency (wrt to 16 cores) vs number of cores, to help see the effect of problem size and iso-efficiency}
\label{3depths}
\vspace{-0.3in}
\end{figure}
The results are shown in Figure~\ref{3depths}.  In the easiest case (depth 8),
there is not enough additional parallel work to justify the jump from 256 to
1,024 processors, and performance actually degrades.  Aside from this anomaly,
however, we see consistent speedups as we increase the number of processors. 
The plot on the bottom helps to illustrate the effect of problem size on parallel efficiency (defined as Speedup/P) and helps understand the iso-efficiency of the problem. Iso-efficiency~\cite{b:grama-isoefficiency} means how should the problem size be increased with increasing number of processors to maintain same level of parallel efficiency. As the size of the problem increases, the amount of parallelizable work available also increases, and hence parallel efficiency is higher for larger problem sizes and constant number of processors (Figure~\ref{3depths} bottom). As the number of processors increases, parallel efficiency decreases for constant problem size. Figure~\ref{3depths} suggests that in parallel information set generation, to maintain constant parallel efficiency, problem size (search depth) should be roughly increased by two when the number of processor in increased by 4 times.

@@

\centering
%\begin{tabular}{rrrrr}
\begin{tabular}{ccccc}
PE/G & 2 & 4 & 6 & 8 \\
1 & {\bf 177$|$1} & 178$|$0.99 & 1844$|$0.96 & 493$|$0.36 \\
4 & 71$|$2.48 & 53$|$3.34 & {\bf 48$|$3.63} & 75$|$2.35 \\
16 & 34$|$5.19 & 19 $|$9.27& {\bf 12$|$13.8} & 16$|$10.7 \\
64 & 24 $|$7.25 & 7 $|$23.6& {\bf 3.7$|$47.39} & 4.1$|$43.1 \\
256 & 24 $|$ 7.1& 4.4 $|$40.1 & 1.2 $|$145& {\bf 1.0$|$172} \\
1024 & 26 $|$6.61 & 3.8 $|$46.5& 0.64 $|$276& {\bf 0.40$|$443}
\end{tabular}

@@

\label{tab:prob5}
\vspace{-0.1in}
\end{table}

@@

\begin{figure}[t]
 \centering \includegraphics[scale=1]{plots/new.pdf} %
\caption{Performance results for Information Set Generation for two games of Kriegspiel, position ten}
\label{prob4} 
\end{figure}

Note also that information set generation is just one subroutine needed to
reason about partially observable games.  Once plausible game histories have
been found, an agent may simulate some sequences of moves up to previous moves
made by opponents and then recursively call the ISG procedure from the
perspective of the opponent.  This results in many parallel applications of
ISG. Furthermore, given a node in an information set, it is common for an agent
to perform forward search on that node (e.g., using MCTS).  All of these
forward search operation can be performed independently in parallel. With these
two additional levels of parallelism on top of the information set generation
problem, it is likely that the problem of playing partially observable
games could scale efficiently to millions of processors or more.
\begin{table}[ht]
\caption{Best Speedups reported for game tree applications.}
\centering
\begin{tabular}{lrrc}
{\bf Authors} & {\bf Game} & {\bf Year} & {\bf Best Speedup (cores)} \\
Ferguson \& Korf & Othello & 1988 & 12(32) \\
Felton \& Otto & Chess & 1988 & 100(256) \\
Richards \etal & Kriegspiel & 2012 & 178(256) \\
Richards \etal & Kriegspiel & 2012 & {\bf 536(1024)} \\
\end{tabular}
\label{bestspeedups}
\end{table}
\section{Related Work}
\label{Related}
An overview of the theory of imperfect information games and other basic topics
in game theory can be found in~\cite{kuhn03lectures} and~\cite{kuhn97classics}.
A technical explanation of the concept of information sets can be found
in~\cite{gilpin09algorithms}.  Koller \etal, showed that a Nash equilibrium can
be found in time that is polynomial in the number of nodes in the game
tree~\cite{koller94fast}.  In fact, they found that it took about as much time
to generate the full game tree as it did to actually solve it.  Unfortunately,
this is feasible only for small games.  In larger games where theoretically
optimal strategies cannot be computed efficiently, computer systems tend to
utilize the concept of information sets or belief states in some form or
another.  A common strategy is to estimate the value of a move by averaging the
estimated value of the associated descendant for each node in the current
information set (or a random sample from the set).  While this sort of
``averaging over clairvoyance'' {\em can} lead to very poor decisions, it is
nevertheless a reasonable strategy in some games, such as
Scrabble~\cite{sheppard02world}.

@@

>>>>

%\documentclass[times, 10pt,twocolumn]{article}
%\documentclass[conference]{IEEEtran}
\documentclass[10pt, conference, compsocconf]{IEEEtran} 
%\usepackage{latex8}
%\usepackage{times}
\usepackage{url,hyperref}
\usepackage{subfigure}
\usepackage{tikz}
\usepackage{examplep}
\usepackage{moreverb}
\usepackage{algorithm}
\usepackage{boxedminipage}
\usepackage[noend]{algpseudocode}
\usepackage{listings}
\usepackage[cmex10]{amsmath}
\usepackage{amsfonts,latexsym,graphicx}
\usepackage[tableposition=top,font=bf]{caption}
%\usepackage[tableposition=top,font=bf]{caption}

@@

\newcommand{\etal}{{\em et al.}\ }
\newcommand{\assign}{\leftarrow}
\newcommand{\eps}{\epsilon}
%\renewcommand{\thefootnote}{\fnsymbol{footnote}}
%\documentstyle[times,art10,twocolumn,latex8]{article}

@@

\begin{figure}
\begin{center}
\begin{verbatim}
1. d4 {(:)}
   a5 {(:)}
2. Bg5 {(:)}
   b6 {(:)}
3. Nc3 {(:Bd8)}
   c5 {(:)}
4. d5 {(Tc5:Bd8)}
   Na6 {(:)}
5. d6 {(:)}
   f6 {(Td6:)}
6. e4 {(:)}
   fxg5 {(Td6,Tg5,Xg5:)}
7. Bc4 {(:)}
   exd6 {(Td6,Xd6:)}
8. Qd5 {(:)}
   Qc7 {(:)}
9. Qf7+ {(:)} 
   Kd8 {:g6,Kxf7,Ke7}
10. Qxf8++ {(1-0:)}
\end{verbatim}
\end{center}
\caption{Listing of a Kriegspiel game resulting in checkmate by white after 10 moves. The final, legal move for each player is shown followed by the list of moderator declarations and all moves that were attempted but disallowed before the legal move was accepted.}
\label{listing}
\vspace{-0.1in}
\end{figure}

@@

Figure~\ref{fig:figure1} shows the true state of the game after nine moves by
each player.  Figure~\ref{fig:figure2} shows White's perspective of the game at
that point. At move 9, Black attempts to block a
potential threat along the diagonal by moving a pawn to g6, then attempts to
escape or capture a threat at f7.  When both of these attempts fail, Black can
infer that his king is being threatened by a protected bishop or queen at f7.
The attempt to move to e7 is Black's way of finding out whether the threatening
piece is a queen or bishop.  When this move fails, black knows it is White's
queen and ultimately retreats to d8.  Note that the failed moves by White on
turns 3 and 4 are a result of being blocked by the pawn at e7.

The listing in Figure~\ref{listing} corresponds to the referee's view of the game and would be useful for
post-game analysis or commentary.  During the game, players would not have access to the full information.  The
transcript from White's perspective is shown in Figure~\ref{filteredlisting}.  Note that Black's actual moves have been
replaced with ``??'' and lists of Black's illegal moves have been replaced with the {\em number} of illegal attempted
moves.  For example, on move 9, Black attempted three illegal moves.
\begin{figure}
%\begin{minipage}[b]{0.35\linewidth}
%\centering
%\small
\begin{verbatim}
1. d4 {(:)}
   ?? {(:0)}
2. Bg5 {(:)}
   ?? {(:0)}
3. Nc3 {(:Bd8)}
   ??  {(:0)}
4. d5 {(Tc5:Bd8)}
   ??  {(:0)}
5. d6 {(:)}
   ?? {(Td6:0)}
6. e4 {(:)}
   fxg5 {(Td6,Tg5,Xg5:0)}
7. Bc4 {(:)}
   ?? {(Td6,Xd6:0)}
8. Qd5 {(:)}
   ?? {(:0)}
9. Qf7+ {(:)} 
   ?? {:3}
10. Qxf8++ {(1-0:)}
\end{verbatim}
%\end{verbatim}
%\end{minipage}
%\begin{minipage}[b]{0.25\linewidth}
%\small
%\centering
%\begin{verbatim}
%\end{minipage}
%\vspace{-0.1in}
\caption{Observations received by White over the course of the game.  White does not know Black's actual moves but hears moderator announcements related to captures, pawn tries, and the number of attempted moves by Black that fail.}
\label{filteredlisting}
\vspace{-0.2in}
\end{figure}

@@

Table~\ref{speedups} shows the performance results for our algorithm on a
randomly selected problem instance in which the size of the information set was
about 4.5 million sequences.  This instance is not necessarily representative of all
games or states of the game, but it was not cherry-picked in any way. The table
shows that our implementation achieves good parallel scaling. The speedups are nearly
linear for up to 32 processors, and the speedups continue to increase even up
to 1,024 processors. In this particular instance, the sequential algorithm
takes almost 25 minutes to run.  However, when parallelized on 1,024
processors --- perfectly reasonable resource usage on modern hardware --- the
search takes less than three seconds.  This is a speedup of 576 times! In a
tournament where each player is allotted an average of five minutes per move,
an ISG operation that requires 25 minutes or more would be prohibitively
expensive.  On the other hand, if the operation can be done in under three
seconds, ISG can be used as a subroutine in the decision-making process
(see~\cite{parker05game}).

@@

\caption{Speedups for a randomly selected problem instance on 1--1,024
processors.  The input observations were from the 10th ply of the game tree and
the size of the information set in this instance is 4,495,121 sequences.}
\centering
\begin{tabular}{ccc}
{\bf Procs}	&	{\bf Time (s)} 	&	{\bf Speedup}\\
1	&	1524	&	1\\
2	&	812	&	1.88\\
4	&	407	&	3.74\\
8	&	208	&	7.32\\
16	&	102	&	14.8\\
32	&	53.2	&	28.7\\
64	&	28.0	&	54.5\\
128	&	14.9	&	102\\
256	&	7.70	&	197\\
512	&	4.59	&	331\\
1024	&	2.64	&	576\\
\end{tabular}

@@

We next study the variation in the efficiency of the parallelization of the
algorithm as size of the problem increases.  Here, we varied the number
of plies in the test instance from 8 to 12. We used the work stealing
load-balancing strategy.  
\begin{figure}[ht] 
\centering
\begin{minipage}{0.7\linewidth}
\includegraphics[width=\columnwidth]{plots/3depths.pdf}
\end{minipage}
\centering
\begin{minipage}{0.7\linewidth}
\includegraphics[width=\columnwidth]{plots/3depthspe.pdf}
\end{minipage}
\caption{(Top) Execution time for three different sizes of problem as the number of cores used increases.
(Bottom) Parallel Efficiency (wrt to 16 cores) vs number of cores, to help see the effect of problem size and iso-efficiency.}
\label{3depths}
\vspace{-0.3in}
\end{figure}
The results are shown in Figure~\ref{3depths}.  In the easiest case (depth 8),
there is not enough additional parallel work to justify the jump from 256 to
1,024 processors, and performance actually degrades.  Aside from this anomaly,
however, we see consistent speedups as we increase the number of processors. 
The plot on the bottom helps to illustrate the effect of problem size on parallel efficiency (defined as Speedup/P) and helps understand the iso-efficiency of the problem. Iso-efficiency~\cite{b:grama-isoefficiency} means how should the problem size be increased with increasing number of processors to maintain same level of parallel efficiency. As the size of the problem increases, the amount of parallelizable work available also increases, and hence parallel efficiency is higher for larger problem sizes and constant number of processors (Figure~\ref{3depths} bottom). As the number of processors increases, parallel efficiency decreases for constant problem size. Figure~\ref{3depths} suggests that in parallel information set generation, to maintain constant parallel efficiency, problem size (search depth) should be roughly increased by two when the number of processor in increased by 4 times.

@@

\centering
%\begin{tabular}{rrrrr}
\begin{tabular}{ccccc}
PE/G & 2 & 4 & 6 & 8 \\
1 & {\bf 177$|$1} & 178$|$0.99 & 184$|$0.96 & 493$|$0.36 \\
4 & 71$|$2.48 & 53$|$3.34 & {\bf 48$|$3.63} & 75$|$2.35 \\
16 & 34$|$5.19 & 19 $|$9.27& {\bf 12$|$13.8} & 16$|$10.7 \\
64 & 24 $|$7.25 & 7 $|$23.6& {\bf 3.7$|$47.39} & 4.1$|$43.1 \\
256 & 24 $|$ 7.1& 4.4 $|$40.1 & 1.2 $|$145& {\bf 1.0$|$172} \\
1024 & 26 $|$6.61 & 3.8 $|$46.5& 0.64 $|$276& {\bf 0.40$|$443}
\end{tabular}

@@

\label{tab:prob5}
\vspace{-0.2in}
\end{table}

@@

\begin{figure}[t]
 \centering \includegraphics[scale=1]{plots/new.pdf} %
\caption{Performance results for Information Set Generation for two games of Kriegspiel, position ten}
\label{prob4}
\vspace{-0.2in} 
\end{figure}

Note also that information set generation is just one subroutine needed to
reason about partially observable games.  Once plausible game histories have
been found, an agent may simulate some sequences of moves up to previous moves
made by opponents and then recursively call the ISG procedure from the
perspective of the opponent.  This results in many parallel applications of
ISG. Furthermore, given a node in an information set, it is common for an agent
to perform forward search on that node (e.g., using MCTS).  All of these
forward search operation can be performed independently in parallel. With these
two additional levels of parallelism on top of the information set generation
problem, it is likely that the problem of playing partially observable
games could scale efficiently to millions of processors or more.
\begin{table}[ht]
\caption{Best Speedups reported for game tree applications.}
\centering
\begin{tabular}{lrrc}
{\bf Authors} & {\bf Game} & {\bf Year} & {\bf Best Speedup (cores)} \\
Ferguson \& Korf & Othello & 1988 & 12(32) \\
Felton \& Otto & Chess & 1988 & 100(256) \\
Richards \etal & Kriegspiel & 2012 & 178(256) \\
Richards \etal & Kriegspiel & 2012 & {\bf 536(1024)} \\
\end{tabular}
\label{bestspeedups}
\vspace{-0.2in} 
\end{table}
\section{Related Work}
\label{Related}
An overview of the theory of imperfect information games and other basic topics
in game theory can be found in~\cite{kuhn03lectures} and~\cite{kuhn97classics}.
A technical explanation of the concept of information sets can be found
in~\cite{gilpin09algorithms}.  Koller \etal, showed that a Nash equilibrium can
be found in time that is polynomial in the number of nodes in the game
tree~\cite{koller94fast}.  In fact, they found that it took about as much time
to generate the full game tree as it did to actually solve it.  Unfortunately,
this is feasible only for small games.  In larger games where theoretically
optimal strategies cannot be computed efficiently, computer systems tend to
utilize the concept of information sets or belief states in some form or
another.  A common strategy is to estimate the value of a move by averaging the
estimated value of the associated descendant for each node in the current
information set (or a random sample from the set).  While this sort of
``averaging over clairvoyance'' {\em can} lead to very poor decisions, it is
nevertheless a reasonable strategy in some games, such as
Scrabble~\cite{sheppard02world}.

@@

\section*{Acknowledgments}

Authors were supported by grants 
NSF-0904844, DOE DE-SC0001845. We used machine resources on the
Ranger cluster (TACC), under TeraGrid allocation grant TG-ASC050039N supported
by NSF. First and second author contributed equally. 
